Generalized Quasi-linear Approximation and Non-normality in Taylor-Couette Spiral Turbulence

### 2022-10-11 paper/paper.tex
Updates from Overleaf

\section{Introduction}
\label{sec:intro}
%Something about Taylor (1923); GQL.
Since Taylor's classic 1923 paper\cite{1923RSPTA.223..289T}, fluid between two independently rotating cylinders has become a rich laboratory for the understanding of instability, transition to turbulence, and turbulence itself. Indeed, the interplay between experiment, computation, and theoretical understanding evinced in Taylor's paper remains a model for the study of fluid dynamics.

Some of the most common forms of turbulence in nature are anisotropic, with a symmetry broken by rotation, magnetic fields, or boundary effects. In these situations, significant theoretical progress can be made by considering \emph{quasi-linear} (QL) models, in which one retains only those non-linearities that interact with the mean flow. Recently, the generalised quasi-linear approximation (GQL) has been developed to systematically expand the QL idea to include sets of non-linear interactions between two \emph{sets} of modes, low and high, respectively (see \cite{2022arXiv220505513M} for a recent review). GQL has been shown to perform significantly better in reproducing direct numerical simulations (DNS) than QL models across a wide range of paradigm flows including forced zonal jet formation \cite{2016PhRvL.116u4501M}, rotating plane Couette flow \cite{2017JFM...810..412T}, rapidly rotating convection \cite{2018RSPSA.47480422T}, and channel flow\cite{2019Kellam}.

Here, we consider Taylor-Couette (TC) flows at moderate inner and outer Reynolds numbers, $|\Reyn_{i,o}| < 3500$. This is an interesting test bed for GQL because it features numerous intricate patterns that are maintained by non-linear interactions beyond the cascade, local in coefficient space, found at high $\Reyn$. 
One of the key feature of TC flow is that it has non-linear and linear instabilities in different regions of parameter space. In the linearly stable regions, transient growth has been observed in TC flow\cite{2002PhFl...14.3475H, 2002PhFl...14.1655M}. 
We can write the Navier-Stokes equations in schematic form as
\begin{equation}
    \label{eqn:NS-schematic}
    \frac{\partial \mathbf{q}}{\partial t} + \Lop \mathbf{q} = \mathcal{N}(\mathbf{q}),
\end{equation}
where $\Lop$ is a linear operator and $\mathcal{N}$ is a non-linear function of the state vector $\mathbf{q} = \left[ \mathbf{u},  p\right]^T$, where $\mathbf{u}$ is the velocity field and $p$ the pressure.
As noted by \cite{2002PhFl...14.3475H}, the non-normality of the linear operator $\Lop$ for TC flow is a function of its control parameters.
That is, its commutator with its adjoint
\begin{equation}
    [\Lop^\dagger, \Lop] = f(\mu, \eta, \Reyn_i, \Gamma)
\end{equation}
can be zero or non-zero. 
This provides a very useful arena to test the validity of GQL in a system that admits a varying degree of non-normality.

@@

We begin in section~\ref{sec:methods} by detailing our equations, non-dimensionalisation, and numerical methods. Section~\ref{sec:linear} reviews both modal and non-modal linear analyses of spiral turbulence. In section~\ref{sec:nonlinear}, we describe the main results of the paper, the comparison of DNS, QL, and GQL for supercritical and subcritical spiral turbulence. Finally, we offer discussion and concluding results in section~\ref{sec:conclusion}.
%Recently, a subcritical transition to turbulence via interpenetrating spirals has been reported \cite{2020JFM...892A..12C}. This work has a small aspect ratio $\Gamma = 5.26$, which delays the onset of linear instability to considerably higher $\Reyn_1$ for a given $\Reyn_2$.

\section{Methods}
\label{sec:methods}
We solve the incompressible Navier-Stokes equations with unit density for $\mathbf{u}' = \mathbf{u} - \mathbf{U}$,
where $\mathbf{U} = A r + B/r \hat{\mathbf{e}}_\phi$ is circular Couette flow (CCF),
\begin{align}
\label{eqn:NS}
      \frac{\partial \mathbf{u'}}{\partial t} + \mathbf{U} \cdot \nabla \mathbf{u'} + \mathbf{u'} \cdot \nabla \mathbf{U} + \mathbf{u'} \cdot \nabla \mathbf{u'} &= -\nabla p + \nu \nabla^2\mathbf{u'}\\
      \nabla \cdot \mathbf{u'} &= 0.
\end{align}
In order to implement the GQL approximation, we introduce a projection operator to divide $\mathbf{u'}$ into two subsets,
\begin{align}
    \mathbf{u'}_l = \left< \mathbf{u'}\right>_l &= \sum_{m = -\Lambda_\theta}^{\Lambda_\theta} \sum_{k = -\Lambda_z}^{\Lambda_z} \mathbf{\hat{u}'}_{mk}(r) e^{i (m \theta + 2\pi k z/L_z)}\\
    \mathbf{u'}_h = \mathbf{u'} - \mathbf{u'}_l.
\end{align}
GQL can then be written as 
\begin{align}
\label{eqn:GQL}
      \frac{\partial \mathbf{u'}}{\partial t} + \mathbf{U} \cdot \nabla \mathbf{u'} + \mathbf{u'} \cdot \nabla \mathbf{U} +\nabla p - \nu \nabla^2\mathbf{u'} = &- \left<\mathbf{u'}_l \cdot \nabla \mathbf{u'}_l +\mathbf{u'}_h \cdot \nabla \mathbf{u'}_h\right>_l\\
      & +\left<\mathbf{u'}_l \cdot \nabla \mathbf{u'}_h + \mathbf{u'}_h \cdot \nabla \mathbf{u'}_l \right>_h\\
      \nabla \cdot \mathbf{u'} = & 0,
\end{align}
where we have written all linear terms on the left hand side. This form is equivalent to those given in prior works (for example \cite{2017JFM...810..412T}), where separate equations are written for the low and high modes, respectively.
For all runs in this paper, $\Lambda = \Lambda_z = \Lambda_\theta$, though we note it would be interesting to consider models quasilinear in $\theta$ (i.e. $\Lambda_\theta = 0$) with varying $\Lambda_z$. These correspond to the \emph{restricted non-linear} models used with success in shear flow turbulence \cite{2014PhFl...26j5112T, 2015PhFl...27j5104T, 2015PhFl...27a1702B}.

We non-dimensionalise lengths in terms of the gap width $d = R_2 - R_1$ and velocities in terms of the inner cylinder velocity $U_i = R_i \Omega_i$. Taylor-Couette flow can be parameterised in terms of four parameters, the inner cylinder Reynolds number $\Reyn_i$, the ratio of inner to outer radii $\eta \equiv R_i/R_o$, the ratio of outer to inner rotation rates $\mu \equiv \Omega_o/\Omega_i$, and the aspect ratio of the cylinder $\Gamma \equiv L_z/d$. In our computations, the $z$ direction is periodic and thus $L_z$ is the periodicity length. We will also refer to the two Reynolds numbers $\Reyn_i$, $\Reyn_o$, related by
\begin{equation}
    \frac{\Reyn_o}{\Reyn_i} = \frac{\mu}{\eta}.
\end{equation}
It is also quite useful to characterise times in terms of the viscous diffusion time $\tau_\nu \equiv d^2/\nu$. In our units, $\tau_\nu = \Reyn_i$.

We use the pseudo-spectral Dedalus framework (version 2) \cite{2020PhRvR...2b3068B} for all linear and non-linear calculations. We discretize the three velocity components and pressure onto a basis of Chebyshev polynomials in the radial direction, while $\theta$ and $z$ use Fourier bases.
To evolve the solution forward in time, we use the implicit-explicit Runga-Kutta RK443 time stepper; linear terms are implicitly timestepped while non-linear terms are explicit.
Our initial value solver has been extensively tested for TC flow against the linear and non-linear test cases in \cite{1984JFM...146...45M, 1984JFM...146...65M} including linear growth rates, non-linear wave speeds, and torque measurements; our spiral turbulence DNS results also agree with those in \cite{2009PhRvE..80d6315M,2009PhRvE..80f7301D,2011JFM...668..150D}. 
Resolutions and geometric parameters for all runs are given in table~\ref{tab:simulations}.
\begin{table}[!h]
\caption{Simulations}
\label{tab:simulations}
\begin{tabular}{rlllllllll}
\hline
$\Reyn_i$ & $\Reyn_o$ & $\eta$ & $\Gamma$  & $N_r$ & $N_\theta$ & $N_z$& $\Lambda_\theta$, $\Lambda_z$ & Notes\\
\hline
%A & 750  & 0 & 0.875  & 3 & 32 & 64 & 64 & 0, 1, 3, 5 & WVF\\
%B & 1000 & 0 & 0.875 & 3 & 32 & 256 & 64 & 0, 1, 3, 5 & MWF\\
%C & 2000 & 0 & 0.875 & 3 & 32 & 512 & 64 & 0, 1, 3, 5 & TTV\\
%\hline
900 & -3398 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5, 10 & Supercritical SPT\\
700 & -3398 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5, 10, 20, 30 & Subcritical SPT\\
%\hline
%F & 640 & -1359 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5 & Supercritical SPT\\
%G & 480 & -1359 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 5 & Supercritical IPS\\
\end{tabular}
\vspace*{-4pt}
\end{table}%%%End of the table
We have confirmed our simulations are resolved by ensuring that the highest wavenumber components have spectral power at least $10^5$ smaller than the most energetic scale.

@@

\section{Linear Stability}
\label{sec:linear}
In order to determine the linear stability bound, we find the minimum $\Reyn_i$ for a given $\Reyn_o$ by solving the eigenvalue problem for equations~\ref{eqn:NS}. We note that for all $\Reyn_o$ plotted in figure~\ref{fig:LSB}, most unstable mode is never axisymmetric. At $\Reyn_o=-3398$, where our non-linear calculations are run, the most unstable mode has $m = 5$ at onset. Figure~\ref{fig:LSB} also shows the hysteresis boundaries given in references \cite{1986JFM...164..155A} and \cite{1965JFM....21..385C}. In both cases, original tabular data was not available, so we extracted the data from figures 16 and 2a, respectively using WebPlotDigitizer \cite{Rohatgi2020}. Note the error bars from Coles (1965) \cite{1965JFM....21..385C} give the top and bottom of the markers from the original paper. Notably, even with this uncertainty, the two studies do not agree on the lower boundary for hysteresis despite similar experimental geometries. 
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figs/reo_rei_lsb.pdf}
    \caption{A summary of spiral turbulence runs in the $\Reyn_o-\Reyn_i$ plane showing linear stability boundary. Solid circles shows points where we have conducted GQL analysis; crosses represent DNS steps along the path to subcritical behaviour. The curves labelled Andereck et al. (1986) and Coles (1965) give the boundaries of spiral turbulence hysteresis reported in those two papers.}
    \label{fig:LSB}
\end{figure}
The spiral turbulence regime provides an ideal opportunity to test GQL in an environment with both spatiotemporal patterns and a tunable bifurcation: by choosing $\Reyn_i$ and $\Reyn_o$ appropriately, both subcritical and supercritical manifestations of spiral turbulence can be selected.

Following pioneering numerical work \cite{2009PhRvE..79c6309M, 2009PhRvE..80d6315M}, we set the outer cylinder rotation rate to $\Reyn_o = -3398$ and
%for the subcritical path and $\Reyn_o = -1359$ for the supercritical path and 
start a series of DNS runs at a $\Reyn_i$ high enough to trigger spiral turbulence driven by linear instability from a low-amplitude, random inital condition satisfying $\nabla \cdot \mathbf{u} = 0$.
From these two seed runs, we decrease $\Reyn_i$ in steps shown in figure~\ref{fig:LSB} where each point represents a DNS or GQL simulation run for one viscous time. The solid circles in figure~\ref{fig:LSB} represent runs where we performed GQL analyses.

In order to demonstrate the effects non-normality plays, figure~\ref{fig:pseudospectra} shows both the modal spectrum (in blue points) and the $\epsilon-$pseudospectrum for the supercritical and subcritical $\Reyn_i$. The contour value $\epsilon^{-1}$ represents the set of values 
\begin{equation}
\label{eqn:pseudospectrum}
    \lVert (\sigma \mathbf{I} - \Lop)^{-1} \rVert \geq \epsilon^{-1}
\end{equation}
for a given complex number $\sigma = \gamma + i\omega$ here representing the eigenvalue corresponding to the temporal evolution of a linear solution, $e^{\sigma t}$ (see \cite{trefethen2005spectra} for an introduction). Equation~\ref{eqn:pseudospectrum} depends on the choice of norm; here we use the standard energy norm. These contours in turn give a measure of the maximum amplification of a linear disturbance: the further negative contours extend into the upper-half plane, the larger a linear disturbance can grow as a result of non-normality, irrespective of the presence of instability. This linear effect is captured entirely by GQL (and QL, for that matter). What role it plays in the development of subcritical spiral turbulence remains an open question, though our results provide some interesting constraints.
\begin{figure}
    \centering
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_900.00_eta_0.8830_mu_-3.3333_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_700.00_eta_0.8830_mu_-4.2857_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \caption{Spectra (blue points) and pseudospectra (contours) for $\Reyn_i = 900$ (left) and $\Reyn_i = 700$ (right) for a mode with $m=6$ and $k_z = 9.68$. The latter is the z wavenumber at the critical $\Reyn_{i,crit} \simeq 778$. $\omega$ is the frequency and $\gamma$ the growth rate at each point in the domain. The stability boundary is marked with a grey line at $\gamma = 0$. The are seven equally spaced contours from $-1.8$ to 0.}
    \label{fig:pseudospectra}
\end{figure}
\section{Results}
\label{sec:nonlinear}
Figure~\ref{fig:urms_tz_rei700} shows a typical DNS snapshot of the RMS velocity perturbation $u'_{rms}$ in the middle of the gap for $\Reyn_i=700$. Despite being well below the linear stability threshold $\Reyn_{i,crit} \simeq 778$, the flow morphology is nearly identical to the supercritical case $\Reyn_i = 900$ (not shown). 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../figs/urms_tz_rei_700_reo-3398_000278.png}
    \caption{Slice of fluctuating RMS velocity in the middle of the radial domain of a DNS at $\Reyn_i = 700$, $\Reyn_o = -3398$. This solution is well below the linear stability threshold at $\Reyn_{i,crit} \simeq 778$.}
    \label{fig:urms_tz_rei700}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}
We have presented a series of simulations demonstrating the performance of GQL at various cutoffs for Taylor-Couette flow in the spiral turbulence regime. GQL is a significant improvement over QL models in representing statistical quantities, particularly the power spectra. However, it does not retain an unambiguous spiral form in the supercritical regime until $\Lambda \simeq 30$; at subcritical $\Reyn_i$, $\Lambda \simeq 60$ is required to see any sustained flows at all. Interestingly, when GQL solutions do sustain non-CCF flows in the subcritical regime, they also produce spiral patterns. Prior work has focused on the utility of GQL in presenting a more accurate \emph{statistical} picture of turbulent flows than QL \cite{2016PhRvL.116u4501M, 2017JFM...810..412T,2018RSPSA.47480422T,2019Kellam}. This focus is particularly important in light of its potential as as means of improving the accuracy of direct statistical simulation (DSS), a method for studying turbulence by computing its statistics directly, rather than individual realisations \cite{2022arXiv220505513M}. 
We have shown here that it is possible to use GQL as a means to investigate the non-linear interactions of greatest importance for maintaining the intricate patterns manifested in Taylor-Couette flow. In future work, we will consider the ability of GQL to maintain exact coherent states identified by Deguchi et al \cite{2014PhRvL.112r4502D} and the dynamics of which were studied recently by Wang and coworkers \cite{2022arXiv220712990W}. Understanding how these interactions work to shape subcritical TC turbulence represents a promising new avenue to understanding how ECS are maintained and how they in turn sustain turbulence below linear stability bounds. In particular, one can begin to understand the relative importance of non-normal effects (represented entirely with GQL) and non-linear ECS.

>>>>

\section{Introduction}
\label{sec:intro}
Since Taylor's classic 1923 paper\cite{1923RSPTA.223..289T}, fluid between two independently rotating cylinders has become a rich laboratory for the understanding of instability, transition to turbulence, and turbulence itself. Indeed, the interplay between experiment, computation, and theoretical understanding evinced in Taylor's paper remains a model for the study of fluid dynamics.

Some of the most common forms of turbulence in nature are anisotropic, with a symmetry broken by rotation, magnetic fields, or boundary effects. In these situations, significant theoretical progress can be made by considering \emph{quasi-linear} (QL) models, in which one retains only those non-linearities that interact with the mean flow. Recently, the generalised quasi-linear approximation (GQL) has been developed to systematically expand the QL idea to include sets of non-linear interactions between two \emph{sets} of modes, low and high, respectively (see \cite{2022arXiv220505513M} for a recent review). GQL has been shown to perform significantly better in reproducing direct numerical simulations (DNS) than QL models across a wide range of paradigm flows including forced zonal jet formation \cite{2016PhRvL.116u4501M}, rotating plane Couette flow \cite{2017JFM...810..412T}, rapidly rotating convection \cite{2018RSPSA.47480422T}, and channel flow\cite{2019Kellam, 2022JFM...936A..33H, 2022JFM...944A..34H}.

Here, we consider Taylor-Couette (TC) flows at moderate inner and outer Reynolds numbers, $|\Reyn_{i,o}| < 3500$. This is an interesting test bed for GQL because it features numerous intricate patterns that are maintained by non-linear interactions beyond the cascade, local in coefficient space, found at high $\Reyn$. 
One of the key features of TC flow is that it has non-linear and linear instabilities in different regions of parameter space. In the linearly stable regions, transient growth has been observed in TC flow\cite{2002PhFl...14.3475H, 2002PhFl...14.1655M}. 
We can write the Navier-Stokes equations in schematic form as
\begin{equation}
    \label{eqn:NS-schematic}
    \frac{\partial \mathbf{q}}{\partial t} + \Lop \mathbf{q} = \mathcal{N}(\mathbf{q}),
\end{equation}
where $\Lop$ is a linear operator and $\mathcal{N}$ is a non-linear function of the state vector $\mathbf{q} = \left[ \mathbf{u},  p\right]^T$, where $\mathbf{u}$ is the velocity field and $p$ the pressure.
As noted by \cite{2002PhFl...14.3475H}, the non-normality of the linear operator $\Lop$ for TC flow is a function of its control parameters.
That is, its commutator with its adjoint
\begin{equation}
    [\Lop^\dagger, \Lop] = f(\mu, \eta, \Reyn_i, \Gamma)
\end{equation}
can be zero or non-zero, and $f$ is a function of the four parameters of Taylor-Couette flow: the inner cylinder Reynolds number $\Reyn_i$, the ratio of inner to outer radii $\eta \equiv R_i/R_o$, the ratio of outer to inner rotation rates $\mu \equiv \Omega_o/\Omega_i$, and the aspect ratio of the cylinder $\Gamma \equiv L_z/d$. 
This provides a very useful arena to test the validity of GQL in a system that admits a varying degree of non-normality.

@@

We begin in section~\ref{sec:methods} by detailing our equations, non-dimensionalisation, and numerical methods. Section~\ref{sec:linear} reviews both modal and non-modal linear analyses of spiral turbulence. In section~\ref{sec:nonlinear}, we describe the main results of the paper, the comparison of DNS, QL, and GQL for supercritical and subcritical spiral turbulence. Finally, we offer discussion and concluding remarks in section~\ref{sec:conclusion}.
%Recently, a subcritical transition to turbulence via interpenetrating spirals has been reported \cite{2020JFM...892A..12C}. This work has a small aspect ratio $\Gamma = 5.26$, which delays the onset of linear instability to considerably higher $\Reyn_1$ for a given $\Reyn_2$.

\section{Methods}
\label{sec:methods}
We solve the incompressible Navier-Stokes equations with unit density for $\mathbf{u}' = \mathbf{u} - \mathbf{U}$,
where $\mathbf{U} = A r + B/r \hat{\mathbf{e}}_\phi$ is circular Couette flow (CCF) and $A$ and $B$ are the standard constants (see \cite{Chandra1961}, chapter 7),
\begin{align}
\label{eqn:NS}
      \frac{\partial \mathbf{u'}}{\partial t} + \mathbf{U} \cdot \nabla \mathbf{u'} + \mathbf{u'} \cdot \nabla \mathbf{U} + \mathbf{u'} \cdot \nabla \mathbf{u'} &= -\nabla p + \nu \nabla^2\mathbf{u'}\\
      \nabla \cdot \mathbf{u'} &= 0,
\end{align}
where $p$ is the pressure and $\nu$ is the kinematic viscosity.
We solve equations~\ref{eqn:NS} without modification for the DNS runs.
To briefly review, GQL extends the quaslinear approximation by dividing
 $\mathbf{u'}$ into two subsets,
\begin{align}
    \mathbf{u'}_l = \left< \mathbf{u'}\right>_l &= \sum_{m = -\Lambda_\theta}^{\Lambda_\theta} \sum_{k = -\Lambda_z}^{\Lambda_z} \mathbf{\hat{u}'}_{mk}(r) e^{i (m \theta + 2\pi k z/L_z)}\\
    \mathbf{u'}_h = \mathbf{u'} - \mathbf{u'}_l,
\end{align}
which we refer to as the ``low'' and ``high'' modes, respectively.
$\Lambda_z$ and $\Lambda_\theta$ are adjustable parameters that represent the \emph{cutoff} in wavenumber space delineating low from high. 
The non-linearity, when expressed in spectral space is a triadic interaction between wavenumbers. GQL includes all such interactions that fall into three classes: low-low $\to$ low, high-high $\to$ low, and low-high $\to$ high, and it discards all others.
This is a closed system that respects all conservation laws of the original equations \cite{2016PhRvL.116u4501M}.
Most importantly, it represents a systematic closure of increasing fidelity independent of a particular theoretical model for the underlying turbulence.

We introduce a projection operator to perform this division, in which case
GQL can be written as 
\begin{align}
\label{eqn:GQL}
      \frac{\partial \mathbf{u'}}{\partial t} + \mathbf{U} \cdot \nabla \mathbf{u'} + \mathbf{u'} \cdot \nabla \mathbf{U} +\nabla p - \nu \nabla^2\mathbf{u'} = &- \left<\mathbf{u'}_l \cdot \nabla \mathbf{u'}_l +\mathbf{u'}_h \cdot \nabla \mathbf{u'}_h\right>_l\\
      & +\left<\mathbf{u'}_l \cdot \nabla \mathbf{u'}_h + \mathbf{u'}_h \cdot \nabla \mathbf{u'}_l \right>_h\\
      \nabla \cdot \mathbf{u'} = & 0,
\end{align}
where we have written all linear terms on the left hand side. This form is equivalent to those given in prior works (for example \cite{2017JFM...810..412T}), where separate equations are written for the low and high modes, respectively.

We non-dimensionalise lengths in terms of the gap width $d = R_2 - R_1$ and velocities in terms of the inner cylinder velocity $U_i = R_i \Omega_i$. In our computations, the $z$ direction is periodic and thus $L_z$ is the periodicity length. We will also refer to the two Reynolds numbers $\Reyn_i$, $\Reyn_o$, related by
\begin{equation}
    \frac{\Reyn_o}{\Reyn_i} = \frac{\mu}{\eta}.
\end{equation}
It is also quite useful to characterise times in terms of the viscous diffusion time $\tau_\nu \equiv d^2/\nu$. In our units, $\tau_\nu = \Reyn_i$.

In our simulations, the possible spiral patterns are quantised by the periodicity in $z$ and $\theta$. Figure~\ref{fig:urms_tz_rei700} shows that for $\Reyn_i = 700$, the pattern selected has $k_z = m = 1$; $\Reyn_i = 900$ has the same pattern (not shown). Given this pattern, it is unclear a priori how to select the cutoff wavenumbers: unlike the case of plane Couette flow, where a self-sustaining process involving a quasilinear transfer of streamwise waves to spanwise/wall-normal rolls is well established, here we have a turbulent structure that is bounded in both $z$ and $\theta$. 
Given this lack of apparent directionality, for all runs in this paper, $\Lambda = \Lambda_z = \Lambda_\theta$, though we note it would be interesting to consider models with varying $\Lambda_\theta$  but $\Lambda_z = N_z$, where $N_z$ is the DNS resolution. Such models are fully non-linear in $z$ and thus generalise the quasilinear \emph{restricted non-linear} models used with success in shear flow turbulence \cite{2014PhFl...26j5112T, 2015PhFl...27j5104T, 2015PhFl...27a1702B}.

We use the pseudo-spectral Dedalus framework (version 2) \cite{2020PhRvR...2b3068B} for all linear and non-linear calculations. We discretize the three velocity components and pressure onto a basis of Chebyshev polynomials in the radial direction, while $\theta$ and $z$ use Fourier bases.
To evolve the solution forward in time, we use the implicit-explicit Runga-Kutta RK443 time stepper; linear terms are implicitly timestepped while non-linear terms are explicit.
Our initial value solver has been extensively tested for TC flow against the linear and non-linear test cases in \cite{1984JFM...146...45M, 1984JFM...146...65M} including linear growth rates, non-linear wave speeds, and torque measurements; our spiral turbulence DNS results also agree with those in \cite{2009PhRvE..80d6315M,2009PhRvE..80f7301D,2011JFM...668..150D}. 
Resolutions and geometric parameters for all runs are given in table~\ref{tab:simulations}.
\begin{table}[!h]
\caption{Simulations}
\label{tab:simulations}
\begin{tabular}{rlllllllll}
\hline
$\Reyn_i$ & $\Reyn_o$ & $\eta$ & $\Gamma$  & $N_r$ & $N_\theta$ & $N_z$& $\Lambda_\theta$, $\Lambda_z$ & Notes\\
\hline
%A & 750  & 0 & 0.875  & 3 & 32 & 64 & 64 & 0, 1, 3, 5 & WVF\\
%B & 1000 & 0 & 0.875 & 3 & 32 & 256 & 64 & 0, 1, 3, 5 & MWF\\
%C & 2000 & 0 & 0.875 & 3 & 32 & 512 & 64 & 0, 1, 3, 5 & TTV\\
%\hline
900 & -3398 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5, 10 & Supercritical SPT\\
700 & -3398 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5, 10, 20, 30, 60 & Subcritical SPT\\
%\hline
%F & 640 & -1359 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 3, 5 & Supercritical SPT\\
%G & 480 & -1359 & 0.883 & 29.9 & 64 & 512 & 512 & 0, 1, 5 & Supercritical IPS\\
\end{tabular}
\vspace*{-4pt}
\end{table}%%%End of the table
We have confirmed our simulations are resolved by ensuring that the highest wavenumber components have spectral power at least $10^5$ smaller than the most energetic scale.

@@

\section{Linear Stability}
\label{sec:linear}
In order to determine the linear stability bound, we find the minimum $\Reyn_i$ for a given $\Reyn_o$ by solving the eigenvalue problem for equations~\ref{eqn:NS}. We note that for all $\Reyn_o$ plotted in figure~\ref{fig:LSB}, the most unstable mode is never axisymmetric. At $\Reyn_o=-3398$, where our non-linear calculations are run, the most unstable mode has $m = 6$ at onset. Figure~\ref{fig:LSB} also shows the hysteresis boundaries given in references \cite{1986JFM...164..155A} and \cite{1965JFM....21..385C}. In both cases, original tabular data was not available, so we extracted the data from figures 16 and 2a, respectively using WebPlotDigitizer \cite{Rohatgi2020}. The error bars on the Coles (1965) \cite{1965JFM....21..385C} data give the top and bottom of the markers from the original paper. We note in passing that the two studies do not agree on the lower boundary for hysteresis despite similar experimental geometries. Our subcritical experiment is within the more conservative hysteresis boundary given by reference \cite{1986JFM...164..155A}.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figs/reo_rei_lsb.pdf}
    \caption{A summary of spiral turbulence runs in the $\Reyn_o-\Reyn_i$ plane showing linear stability boundary. Solid circles shows points where we have conducted GQL analysis; crosses represent DNS steps along the path to subcritical behaviour. The curves labelled Andereck et al. (1986) and Coles (1965) give the boundaries of spiral turbulence hysteresis reported in those two papers.}
    \label{fig:LSB}
\end{figure}
The spiral turbulence regime provides an ideal opportunity to test GQL in an environment with both spatiotemporal patterns and a tunable bifurcation: by choosing $\Reyn_i$ and $\Reyn_o$ appropriately, both subcritical and supercritical manifestations of spiral turbulence can be selected.

Following pioneering numerical work \cite{2009PhRvE..79c6309M, 2009PhRvE..80d6315M}, we set the outer cylinder rotation rate to $\Reyn_o = -3398$ and
%for the subcritical path and $\Reyn_o = -1359$ for the supercritical path and 
start a DNS run at $\Reyn_i = 900$, high enough to trigger spiral turbulence driven by linear instability from a low-amplitude, random initial condition satisfying $\nabla \cdot \mathbf{u} = 0$.
From this seed run, we decrease $\Reyn_i$ in steps shown in figure~\ref{fig:LSB} where each point represents a DNS or GQL simulation run for one viscous time. The solid circles in figure~\ref{fig:LSB} represent runs where we performed GQL analyses.

In order to demonstrate the effects non-normality plays, figure~\ref{fig:pseudospectra} shows both the modal spectrum (in blue points) and the $\epsilon-$pseudospectrum for the supercritical and subcritical $\Reyn_i$. The contour value $\log_{10} \epsilon^{-1}$ represents the set of values 
\begin{equation}
\label{eqn:pseudospectrum}
    \lVert (z \mathbf{I} - \Lop)^{-1} \rVert \geq \epsilon^{-1}
\end{equation}
for a given complex number $z = \gamma + i\omega$. The spectrum {$\sigma$} are the eigenvalues corresponding to the temporal evolution of a linear solution, $e^{\sigma t}$ (see \cite{trefethen2005spectra} for an introduction). Equation~\ref{eqn:pseudospectrum} depends on the choice of norm; here we use the standard energy norm. These contours in turn give a measure of the maximum amplification of a linear disturbance: the further negative contours extend into the upper-half plane, the larger a linear disturbance can grow as a result of non-normality, irrespective of the presence of instability. This linear effect is captured entirely by GQL (and QL, for that matter). However, the role it plays in the development of subcritical spiral turbulence remains an open question, as the relevant linear operator $\Lop$ in the subcritical case is not that associated with circular Couette flow $\mathbf{U}$. Instead, any transient growth is generated by the saturated mean velocity profile.
\begin{figure}
    \centering
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_900.00_eta_0.8830_mu_-3.3333_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_700.00_eta_0.8830_mu_-4.2857_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \caption{Spectra (blue points) and pseudospectra (contours of $\log_{10}\epsilon^{-1}$) for $\Reyn_i = 900$ (left) and $\Reyn_i = 700$ (right) for a mode with $m=6$ and $k_z = 9.68$. The latter is the z wavenumber at the critical $\Reyn_{i,crit} \simeq 778$. $\omega$ is the frequency and $\gamma$ the growth rate at each point in the domain. The stability boundary is marked with a grey line at $\gamma = 0$. The are seven equally spaced contours from $-1.8$ to 0.}
    \label{fig:pseudospectra}
\end{figure}
\section{Results}
\label{sec:nonlinear}
Figure~\ref{fig:urms_tz_rei700} shows a typical DNS snapshot of the RMS velocity perturbation $u'_{rms}$ in the middle of the gap for $\Reyn_i=700$. Despite being well below the linear stability threshold $\Reyn_{i,crit} \simeq 778$, the flow morphology is nearly identical to the supercritical case $\Reyn_i = 900$ (not shown). 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../figs/urms_tz_rei_700_reo-3398_000278.png}
    \caption{Slice of fluctuating RMS velocity in the middle of the radial domain of a DNS at $\Reyn_i = 700$, $\Reyn_o = -3398$. This solution is well below the linear stability threshold at $\Reyn_{i,crit} \simeq 778$.}
    \label{fig:urms_tz_rei700}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}
We have presented a series of simulations demonstrating the performance of GQL at various cutoffs for Taylor-Couette flow in the spiral turbulence regime. GQL is a significant improvement over QL models in representing statistical quantities, particularly the power spectra. However, it does not retain an unambiguous spiral form in the supercritical regime until $\Lambda \simeq 30$; at subcritical $\Reyn_i$, $\Lambda \simeq 60$ is required to see any sustained flows at all. Interestingly, when GQL solutions do sustain non-CCF flows in the subcritical regime, they also produce spiral patterns. Prior work has focused on the utility of GQL in presenting a more accurate \emph{statistical} picture of turbulent flows than QL \cite{2016PhRvL.116u4501M, 2017JFM...810..412T,2018RSPSA.47480422T,2019Kellam}. This focus is particularly important in light of its potential as a means of improving the accuracy of direct statistical simulation (DSS), a method for studying turbulence by computing its statistics directly, rather than individual realisations \cite{2022arXiv220505513M}. 
We have shown here that it is possible to use GQL as a means to investigate the non-linear interactions of greatest importance for maintaining the intricate patterns manifested in Taylor-Couette flow. In future work, we will consider the ability of GQL to maintain exact coherent states identified by Deguchi et al \cite{2014PhRvL.112r4502D} and the dynamics of which were studied recently by Wang and coworkers \cite{2022arXiv220712990W}. Understanding how these interactions work to shape subcritical TC turbulence represents a promising new avenue to understanding how ECS are maintained and how they in turn sustain turbulence below linear stability bounds. In particular, one can begin to understand the relative importance of non-normal effects (represented entirely with GQL) and non-linear ECS.
